---
title: "Revisiting the SAFE Framework in the Statcast Era: A Modernized Approach to Evalutating MLB Infield Defense"
author:
  - name: Tyler McCord
    corresponding: true
    email: tmccord26@amherst.edu
    affiliations:
      - Amherst College
keywords: Bayesian modeling, Sports analytics, Baseball defense
abstract: |
  High resolution tracking data has transformed player evaluation in Major League Baseball (MLB), enabling high-level analysis of player performance. While public analyses on batting and pitching have advanced rapidly, defensive evaluation has been comparatively underdeveloped. The SAFE (Spatially Adjusted Fielding Evaluation) framework, introduced by @jensen_2009, was the first effort in the public sphere to evaluate defense as a continuous space. We revisit the SAFE framework using modern Statcast data with an emphasis on infield defense, a notable struggle for prior defensive metrics. [Placeholder for results]
date: last-modified
bibliography: references.bib
number-sections: true
---

```{r}
library(tidyverse)
library(kableExtra)
```


## Introduction

The evaluation of batting and pitching in baseball has been at the forefront of sports analytics for decades, mostly due to their discrete nature and the availability of relevant, quantifiable data.
It is relatively simple to measure the outcome of a plate appearance or a pitch, making it easier to develop metrics that accurately reflect player performance in these areas.
In contrast, defensive evaluation has lagged behind due to the continuous, spatiotemporal nature of fielding. 

Still, Major League Baseball (MLB) organizations are faced with important decisions regarding defense, such as positioning players, making defensive substitutions, and evaluating trade-offs between offensive and defensive abilities.
At the end of each season, MLB issues Gold Glove awards to the best defenders at each position, highlighting the importance of defense in the game.

Before the advent of high-resolution player tracking data, teams relied on simple defensive metrics such as fielding percentage, which calculates the percentage of plays a fielder successfully makes, and errors, which count the number of plays that the player does not make that the average fielder would.
However, errors are prone to subjectivity, as they depend primarily on the official scorer's judgement.
These metrics also fail to capture the full scope of a player's defensive contributions, as they do not account for factors such as range, positioning, and the difficulty of plays made.

Statisticians have tried to find ways to quantify the nuances of defense.
In 2003, Mitchel Lichtman introduced the Ultimate Zone Rating (UZR) metric, which attempted to evaluate defense by dividing the field into discrete zones and assigning run values to plays made or not made within those zones (@uzr_fangraphs).
This run-based approach allowed statisticians to understand the stakes of each defensive play.

In 2009, @jensen_2009 introduced the SAFE (Spatially Adjusted Fielding Evaluation) framework, which built upon UZR by using a hierarchical Bayesian model to evaluate defense as a continuous surface.
The SAFE framework uses estimates of player location, ball location, and ball velocity to model the probability of a fielder making a play on a batted ball, allowing for a more nuanced evaluation of defensive performance.
The model combines the probability of a made play with the run consequences of that play to estimate the overall defensive contribution of a player in terms of runs saved or allowed.
The hierarchical Bayesian structure also allows for the sharing of information across players and positions, improving estimates for players with limited data.
However, this model is limited by the accuracy and reliability of the underlying data used to estimate player and ball locations.
These data, provided by Baseball Info Solutions, used hand-annotated video footage to estimate ball location and velocity.
Even then, the starting location of the fielder at a given position was estimated by the authors by using the average location of balls caught by that position.

Notably, the results of @jensen_2009 showed that the autocorrelation of defensive metrics from year to year was quite low for infielders. 
This shortcoming suggests that the original SAFE model performed poorly in evaluating infield defense, relative to outfield defense.

Since the publication of the SAFE framework, MLB has introduced Statcast, a high-resolution player tracking system that uses a combination of radar and camera technology to track the movement of players and the ball in real-time (need to find a citation).
Statcast provides a wealth of data that was previously unavailable, including precise measurements of player and ball locations, velocities, and trajectories.
This data has the potential to revolutionize defensive evaluation in baseball, allowing for more accurate and reliable estimates of defensive performance.

In this paper, we modernize the original SAFE framework for infielders using three years Statcast data (2023-2025).
We propose three models:

1. A reproduction of the original SAFE model using Statcast data.
2. A semi-reproduction of the original SAFE model using updated player-specific positioning data.
3. An improved model that incorporates player-specific positioning data and additional covariates to better capture the nuances of infield defense.

We compare the results of these models to assess the impact of using high-resolution tracking data and updated positioning information on defensive evaluation.
We also evaluate the year-to-year reliability of our models to assess their effectiveness in capturing true defensive ability.

## Data

Our evaluation of infield defense is based on Statcast data from 2023-2025.
Although Statcast data has been publicly available since 2015, we focus on the most recent three years because the infield "shift", a defensive strategy where infielders position themselves in extreme positions based on the batter's hitting tendencies, was banned following the 2022 season.
We believe that narrowing the frame of our analysis to non-shifted seasons will yield more accurate estimates due to more consistent estimates for fielder locations.
We obtain data on batted balls through the `baseballr` package in R, which provides a convenient interface for scraping Statcast data from MLB's public API.
The result is an .Rds file for each year of interest where each observation corresponds to a single batted ball in play (BIP) event.
Further, as an extension of the original SAFE framework, we extract information on individual player positioning before each pitch by using the "Fielder Positioning" page on Baseball Savant (@savant_positioning). 
The location for each infielder on a given play is not publicly available.
However, Baseball Savant provides average positioning data for each player at each position based on all batted balls successfully fielded by that position.
We extract these data by downloading the relevant CSV files from the Baseball Savant website. 

For each batted ball in play, we extract the relevant information needed to identify the fielder responsible for making a play, the location and velocity of the batted ball, and the outcome of the play.

Using these data, we derive the following factors for each batted ball:

- **successful_play**: A binary indicator of whether the fielder successfully made a play on the batted ball (1 = successful play, 0 = unsuccessful play). For ground balls, this is defined as whether the fielder was able to field the ball and record at least one out. For fly balls/line drives, this is defined as whether the fielder was able to catch the ball before it touched the ground.
- **play_made_by**: The position (1B, 2B, SS, 3B) of the fielder who made the play on the batted ball (Other, if play is not successful or play is made by another position).
- **location_x, location_y**: The (x, y) coordinates of the batted ball when it reaches the fielder's location, measured in feet from home plate. The origin (0, 0) is at home plate, with the positive x-axis extending towards first base and the positive y-axis extending towards second base.
- **spray_angle**: Derived from the (x, y) coordinates, this angle represents the direction of the batted ball relative to home plate, measured in degrees. The first base foul line is 45 degrees, second base is 0 degrees, and the third base foul line is -45 degrees. 
- **launch_velocity**: The velocity at which a ball is hit off the bat, measured in miles per hour (mph).
- **out_{pos}**: A binary indicator for each infielder position (1B, 2B, SS, 3B) indicating whether or not the player at that position recorded a successful play on the batted ball.

The resulting dataset contains 16 variables from 339,230 batted balls in play in fair territory from the 2023-2025 seasons. 


```{r}
#| label: tbl-bip-summary
#| tbl-cap: "Summary statistics for batted balls in play from 2023-2025 seasons."


bip_full <- read_rds("statcast_data/bip_clean.rds")

bip_full |>
  group_by(game_year) |>
  summarise(
    total_bip = n(),
    ground_balls = sum(bb_type == "ground_ball", na.rm = TRUE),
    line_drives = sum(bb_type == "line_drive", na.rm = TRUE),
    fly_balls = sum(bb_type == "fly_ball", na.rm = TRUE),
    pop_ups = sum(bb_type == "popup", na.rm = TRUE),
    outs_1b = sum(out_1b == 1, na.rm = TRUE),
    outs_2b = sum(out_2b == 1, na.rm = TRUE),
    outs_ss = sum(out_ss == 1, na.rm = TRUE),
    outs_3b = sum(out_3b == 1, na.rm = TRUE)
  ) |>
  ungroup() |>
  kable(
    booktabs = TRUE,
    col.names = c("Year", "Total BIP", "Ground Balls", "Line Drives", "Fly Balls", "Pop Ups", "Outs at 1B", "Outs at 2B", "Outs at SS", "Outs at 3B"),
    )
```

@tbl-bip-summary shows summary statistics for batted balls in play hit into fair territory from the 2023-2025 seasons.
The proportion of batted balls by type and the number of outs recorded at each infield position are consistent across the three seasons.

```{r}
#| label: tbl-continuous-summary
#| tbl-cap: "Summary statistics for batted ball variables in the batted ball dataset."


continuous_vars <- c("launch_speed", "spray_angle", "location_x", "location_y")

bip_full |>
  select(continuous_vars) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  mutate(variable = factor(variable, levels = continuous_vars,
                           labels = c("Exit Velocity (mph)", "Spray Angle (degrees)", "Location X (ft)", "Location Y (ft)"))) |>
  group_by(variable) |>
  summarize(
    mean = round(mean(value, na.rm = TRUE), 2),
    sd = round(sd(value, na.rm = TRUE), 2),
    min = round(min(value, na.rm = TRUE), 2),
    max = round(max(value, na.rm = TRUE), 2)
  ) |>
  kable(
    col.names = c("Variable", "Mean", "SD", "Min", "Max")
  )
```


```{r}
#| label: fig-fielder-positioning
#| fig-cap: "Average fielder positioning for infielders (1B, 2B, SS, 3B) based on batted balls successfully fielded from 2023-2025 seasons."

safe_positioning <- read_rds("statcast_data/player_positioning/safe_positioning.rds")

ggplot() +
  geom_point(data = bip_full |> filter(play_made_by != "Other"), aes(x = location_x, y = location_y), alpha = 0.1) +
  geom_point(data = safe_positioning, aes(x = pos_x, y = pos_y), color = "red", size = 3) +
  facet_wrap(~ play_made_by) +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )
```

[Might get moved to a Methods section] Following the process outlined in @jensen_2009, we derive the starting location of the fielder at each position by calculating the average (x, y) coordinates of all batted balls successfully fielded by that position. @fig-fielder-positioning shows the calculated average fielder positioning for each infield position based on batted balls successfully fielded from the 2023-2025 seasons.

## Methods

To describe the modernization of the SAFE framework, we start with a technical overview of the original model from @jensen_2009.

The first step is to estimate the probability that a fielder makes a successful play on a batted ball for each combination of year, position, and batted ball type (ground ball, line drive, fly ball/pop up).

### Fly Balls and Line Drives

For fly balls and line drives, a successful play is defined as a play in which the fielder catches the ball before it touches the ground.
Then, we can define each batted ball $j$  as a Bernoulli trial with outcome $S_{ij} = 1$ if player $j$ made a successful play and $S_{ij} = 0$ otherwise.
Then, we define:
$$
S_{ij} = \text{Bernoulli}(p_{ij})
$$
where $p_{ij}$ is the probability that fielder $i$ makes a successful play on batted ball $j$.

We model this probability as a function of the distance that player $i$ must travel to reach the location of batted ball $j$ ($D_{ij}$), the velocity of the batted ball ($V_{ij}$), and an indicator for the direction that the fielder must travel to make the play ($F_{ij} = 1$ for forward, $F_{ij} = 0$ for backward):

$$
p_{ij} = \Phi(\beta_{i0} + \beta_{i1} D_{ij} + \beta_{i2} D_{ij} F_{ij} + \beta_{i3} D_{ij} V_{ij} + \beta_{i4} D_{ij} V_{ij} F_{ij})
$$

where $\Phi$ is the standard normal cumulative distribution function (CDF) and $\beta_{i0}, \beta_{i1}, \beta_{i2}, \beta_{i3}, \beta_{i4}$ are player-specific coefficients that capture the effect of each covariate on the probability of making a successful play.

```{r}

# --------------------- FLY BALL/POP UP MODEL FITTING ---------------------
bip_fb_pu <- bip_full |>
  filter(bb_type %in% c("fly_ball", "popup"))

fit_position_model <- function(bip_data, year, positioning_data, position) {
  # Filter positioning data for the relevant year and position
  pos <- positioning_data |>
    filter(play_made_by == !!position)
  
  if (nrow(pos) != 1) {
    stop("Positioning data must have exactly one row for each year/position combination")
  }
  
  # Merge position coordinates onto play data, filter to year, and compute features
  play_data <- bip_data |>
    filter(game_year == year) |>
    mutate(
      pos_x = pos$pos_x,
      pos_y = pos$pos_y,
      d_pos = sqrt(pos_x^2 + pos_y^2),
      d_ball = sqrt(location_x^2 + location_y^2),
      dist_to_ball = sqrt((location_x - pos_x)^2 + (location_y - pos_y)^2),
      forward = ifelse(d_pos > d_ball, 1, 0)
    )

  # Create the outcome variable name dynamically
  outcome_var <- paste0("out_", str_to_lower(position))
  formula_str <- paste0(
    outcome_var, " ~ dist_to_ball + dist_to_ball:forward + dist_to_ball:launch_speed + dist_to_ball:launch_speed:forward"
  )

  # Pre bayesian model 
  model <- glm(
    formula = as.formula(formula_str),
    data = play_data,
    family = binomial(link = "probit")
  )
  
  return(model)
}

```


```{r}

# --------------------- LINE DRIVE MODEL FITTING ---------------------
bip_fb_pu <- bip_full |>
  filter(bb_type == "line_drive")

fit_position_model <- function(bip_data, year, positioning_data, position) {
  # Filter positioning data for the relevant year and position
  pos <- positioning_data |>
    filter(play_made_by == !!position)
  
  if (nrow(pos) != 1) {
    stop("Positioning data must have exactly one row for each year/position combination")
  }
  
  # Merge position coordinates onto play data, filter to year, and compute features
  play_data <- bip_data |>
    filter(game_year == year) |>
    mutate(
      pos_x = pos$pos_x,
      pos_y = pos$pos_y,
      d_pos = sqrt(pos_x^2 + pos_y^2),
      d_ball = sqrt(location_x^2 + location_y^2),
      dist_to_ball = sqrt((location_x - pos_x)^2 + (location_y - pos_y)^2),
      forward = ifelse(d_pos > d_ball, 1, 0)
    )

  # Create the outcome variable name dynamically
  outcome_var <- paste0("out_", str_to_lower(position))
  formula_str <- paste0(
    outcome_var, " ~ dist_to_ball + dist_to_ball:forward + dist_to_ball:launch_speed + dist_to_ball:launch_speed:forward"
  )

  # Pre bayesian model 
  model <- glm(
    formula = as.formula(formula_str),
    data = play_data,
    family = binomial(link = "probit")
  )
  
  return(model)
}
```


## References {.unnumbered}

:::{#refs}

:::